Fail Faster

### 2025-03-16 paper/paper.tex
rip out mentions of RNG

% \subsubsection{Function call overhead}
\subsection{Choice of RNG}
\jwc{We really need to have explained the effect ordering equivalence aspect of the evaluation before here.}

The core of any PBT generator library is a random number generator: to generate
random values, we need \hg{...}
Different PBT libraries use a wide range of different RNGs.
Following the original Haskell QuickCheck implementation, \bq\ uses the SplitMix
algorithm \cn,
implemented as a (stateful) OCaml library called \texttt{Splittable\_random}. Meanwhile,
ScalaCheck uses the JSF algorithm \cn.
% https://www.pcg-random.org/posts/bob-jenkins-small-prng-passes-practrand.html
% https://burtleburtle.net/bob/rand/smallprng.html

The RNG sits at the heart of the hot path.
Even basic generators, like ones generating a single \texttt{int} or \texttt{float}, may sample from the RNG multiple times,
and generator combinators like \texttt{list} call those basic generators.
For instance, to generate a single list with 3 elements, 
\hg{To generate a single list with 3 elements, \bq's list generator calls the RNG \dots times!}
Because of this, the speed of a single RNG call matters a great deal. Unfortunately,
existing PBT libraries make relatively inefficient choices on this front,
leading to worse bugfinding power than what is possible.

@@

For example, significantly faster RNG algorithms than SplitMix or JSF exist,
such as the Lehmer RNG \cn. In microbenchmarks, the Lehmer generator almost 2x as fast as SplitMix \cn.
% https://github.com/lemire/testingRNG/tree/master
Moreover, PBT libraries could even consider eschewing the requirement that a source of
entropy pass statistical tests like BigCrush\cn.\hg{This needs more discussion}  This is common practice in other areas of testing already:
fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
\cn. Such approaches are also faster than algorithms like SplitMix: bumping a
pointer and reading from memory (which can be pipelined trivially) will always
be faster on modern CPUs than any RNG with data-dependent arithmetic
instructions.\hg{Should we just always be doing this? I think we want to argue
that this is a bridge too far because you still need to generate the buffer
ahead of time and you can run out of randomness?}

\hg{Alternative framing: ``'': \dots''}
Of course, simply arguing that the RNG is on the hot
path does not guarantee that a faster RNG leads to measurably faster
generation; for that, we need an experiment. The most obvious experiment is to
simply swap out the RNG for \bq\ with a totally different, faster one. But, as
discussed previously, generators with different generation orders can be
difficult to compare. To get around this, we exploit a ``natural experiment:
OCaml's \texttt{Splittable\_random} library is slow in a way that
can be improved \emph{without} changing its extensional behavior.
In particular,
due to implementation details related to the OCaml garbage collector,
values of the OCaml type \texttt{int64} are not machine words, but rather
\emph{pointers} to machine words. This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output, which has a significant performance benefit.
By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
performance can be improved just by using a more performant RNG.

\section{Eliminating Abstraction Overhead of Generator DSLs by Staging and Faster RNG}
\jwc{Emphasize that this is an \emph{COMPLETELY EQUIVALENT drop in replacement!} We din't just build a different library with different distributions.}

@@

\subsection{Basic Design}
\label{subsection:basic-design}
Recall that staging a DSL involves changing the
combinators to run at compile time by carefully annotating their types with \texttt{code}s.
Deciding which types \texttt{t} can
instead be \texttt{t code} --- in other words,
figuring out which parts of the DSL can be determined statically (and can be
part of the compile-time stage), and which parts are only known dynamically (and
hence must be \texttt{code}) --- is an art known as ``binding-time analysis'' \cn.

@@

This type, along with basic monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
The monadic interface is given by a return and bind, as usual.
\texttt{Return} is the constant generator, but this time it 
runs at compile time. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it
returns the generator which always generates that value.
\texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one \footnote{Readers familiar with OCaml may notice that \texttt{return : 'a code -> 'a Gen.t} and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t}
do not have the correct types for a monad instance, preventing us from using \texttt{let\%bind} notation. We rectify this issue in Section~\ref{subsection:codecps} by importing some clever ideas from the staging literature.}.
Operationally, bind takes code for the size and seed, and returns code that (1) let-binds a variable \texttt{a} to spliced-in code that runs \texttt{g}, and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size\_c random\_c}
and \texttt{k .<a>. size\_c random\_c} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the RNG.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}
from Section~\ref{subsection:msp}).
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}

@@

We can then redefine our staged generator monad type to be
\texttt{'a Gen.t = int code -> SR.t code -> 'a CodeCPS.t}, as shown in Figure~\ref{fig:codecps-and-final-gen}.
\jwc{any types that were \texttt{'a Gen.t} before are now \texttt{'a code Gen.t}}.
This gives us the best of both worlds. First, we get a monad instance for \texttt{'a Gen.t} with the correct types, which lets us
use the monadic syntactic sugar of our chosen language. Moreover, we also get to maintain the correct
effect ordering: effectful combinators like \texttt{Gen.int} do their \emph{own} let-insertion, ensuring
that a program like \texttt{Gen.bind Gen.int (fun x -> ...)} generates a let-binding for the result of sampling the RNG.
For example, \texttt{bind (int .<0>. .<1.>) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
\texttt{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}. \jwc{Should we do out the whole reduction sequence here? It might be explanatory, but it also might be boring.}

@@

\jwc{Unclear what we should call this section}
As we discussed in Section~\ref{section:motiv}, inefficient RNG choice is
another bottleneck for finding bugs fast. While generator libraries by
and large use sensible RNG algorithms, they are not explicitly chosen
with performance in mind. Indeed, much faster RNGs \cn and fast sources
of entropy \cn exist. \jwc{... more here?}
To demonstrate that a faster RNG can significanty impact bugfinding power, we
use the natural experiment provided by OCaml's inefficient implementation of
SplitMix. By replacing this slow RNG with a faster but extensionally equivalent
implemetnation, we are able to precisely quantify the bugfinding speedup that using a RNG gives across a range of PBT scenarios.

We emphasize that while we believe that the insight that faster RNG translates to faster bug finding
is a cross-langauge one, the specific technical contents of this section are OCaml-specific.
In most other PBT frameworks \cn, the RNG operates on by machine
integers, so this \emph{particular} inefficiency does not exist.

>>>>

% \subsubsection{Function call overhead}
\subsection{Choice of Randomness Library}
\jwc{We really need to have explained the effect ordering equivalence aspect of the evaluation before here.}

The core of any PBT generator library is a source of randomness: to generate
random values of some datatype, we some access to random numbers!
Different PBT libraries use different randomness libraries implementing different algorithms
\footnote{The common term for such an algorithm or library is a ``Random Number
Generator'' (RNG) we will avoid this term and instead say ``randomness library''
to avoid confusing RNG implementations with the PBT gnenerator libraries that
use them.}.
Following the original Haskell QuickCheck implementation, \bq\ uses the SplitMix
algorithm \cn, implemented as a (stateful) OCaml library called \texttt{Splittable\_random}. Meanwhile,
ScalaCheck uses the JSF algorithm \cn.
% https://www.pcg-random.org/posts/bob-jenkins-small-prng-passes-practrand.html
% https://burtleburtle.net/bob/rand/smallprng.html

The randomness library sits at the heart of the hot path.
Even basic generators --- like ones generating a single \texttt{int} or \texttt{float} uniformly within a range --- can sample \emph{unboundedly many}
random numbers, since they usually use versions of rejection sampling to find a value within the range \cn.
Moreover, generator combinators like \texttt{list} usually make $O(n)$ calls to even those basic generators.
Because of this, the speed of a single sample matters a great deal. Unfortunately,
existing PBT libraries make relatively inefficient choices on this front,
leading to worse bugfinding power than what is possible.

@@

For example, significantly faster algorithms than SplitMix or JSF exist,
such as the Lehmer algorithm\cn. In microbenchmarks, the Lehmer algorithm runs almost 2x as fast as SplitMix \cn.
% https://github.com/lemire/testingRNG/tree/master
Moreover, PBT libraries could even consider eschewing the requirement that a source of
entropy pass statistical tests like BigCrush\cn. \hg{This needs more discussion} 
This is common practice in other areas of testing already:
fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
\cn. Such approaches are also faster than algorithms like SplitMix: bumping a
pointer and reading from memory (which can be pipelined trivially) will always
be faster on modern CPUs than any random sampling algorithm with data-dependent arithmetic
instructions.\hg{Should we just always be doing this? I think we want to argue
that this is a bridge too far because you still need to generate the buffer
ahead of time and you can run out of randomness?}

Of course, simply arguing that the randomness library is on the hot
path does not guarantee that a faster sampling leads to measurably faster
generation; for that, we need an experiment. The most obvious experiment is to
simply swap out the randomness library for \bq{} with a totally different, faster one. But, as
discussed previously, generators with different generation orders can be
difficult to compare. To get around this, we exploit a ``natural experiment:
OCaml's \texttt{Splittable\_random} library is slow in a way that
can be improved \emph{without} changing its extensional behavior.
In particular,
due to implementation details related to the OCaml garbage collector,
values of the OCaml type \texttt{int64} are not machine words, but rather
\emph{pointers} to machine words. This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output, which has a significant performance benefit.
By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
performance can be improved just by using a more performant randomness library.

\section{Eliminating Abstraction Overhead of Generator DSLs by Staging and Faster Randomness Libraries}
\jwc{This section needs a much better title...}
\jwc{Emphasize that this is an \emph{COMPLETELY EQUIVALENT drop in replacement!} We din't just build a different library with different distributions.}

@@

\subsection{Design of a Staged Generator DSL}
\label{subsection:basic-design}
Recall that staging a DSL involves changing the
combinators to run at compile time by carefully annotating their types with \texttt{code}s.
Deciding which types \texttt{t} can
instead be \texttt{t code} --- in other words,
figuring out which parts of the DSL can be determined statically (and can be
part of the compile-time stage), and which parts are only known dynamically (and
hence must be \texttt{code}) --- is an art known as ``binding-time analysis'' \cn.

@@

This type, along with basic monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
The monadic interface is given by a return and bind, as usual.
\texttt{Return} is the constant generator, but this time it 
runs at compile time. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it
returns the generator which always generates that value.
\texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one \footnote{Readers familiar with OCaml may notice that \texttt{return : 'a code -> 'a Gen.t} and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t}
do not have the correct types for a monad instance, preventing us from using \texttt{let\%bind} notation. We rectify this issue in Section~\ref{subsection:codecps} by importing some clever ideas from the staging literature.}.
Operationally, bind takes code for the size and seed, and returns code that (1) let-binds a variable \texttt{a} to spliced-in code that runs \texttt{g}, and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size\_c random\_c}
and \texttt{k .<a>. size\_c random\_c} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the randomness library.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}
from Section~\ref{subsection:msp}).
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}

@@

We can then redefine our staged generator monad type to be
\texttt{'a Gen.t = int code -> SR.t code -> 'a CodeCPS.t}, as shown in Figure~\ref{fig:codecps-and-final-gen}.
\jwc{any types that were \texttt{'a Gen.t} before are now \texttt{'a code Gen.t}}.
This gives us the best of both worlds. First, we get a monad instance for \texttt{'a Gen.t} with the correct types, which lets us
use the monadic syntactic sugar of our chosen language. Moreover, we also get to maintain the correct
effect ordering: effectful combinators like \texttt{Gen.int} do their \emph{own} let-insertion, ensuring
that a program like \texttt{Gen.bind Gen.int (fun x -> ...)} generates a let-binding for the result of sampling the randomness library.
For example, \texttt{bind (int .<0>. .<1.>) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
\texttt{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}. \jwc{Should we do out the whole reduction sequence here? It might be explanatory, but it also might be boring.}

@@

\jwc{Unclear what we should call this section}
As we discussed in Section~\ref{section:motiv}, choosing an inefficient randomness library is
another bottleneck for finding bugs fast. While generator libraries by
and large use sensible sources of randomness, they are not explicitly chosen
with performance in mind. Indeed, much faster randomness libraries\cn and fast sources
of entropy \cn exist. \jwc{... more here?}
To demonstrate that faster random sampling can significanty impact bugfinding power, we
use the natural experiment provided by OCaml's inefficient implementation of
SplitMix. By replacing this slow randomness library with a faster but extensionally equivalent
implemetnation, we are able to precisely quantify the bugfinding speedup that using a randomness library gives across a range of PBT scenarios.

We emphasize that while we believe that the insight that faster sampilng translates to faster bug finding
is a cross-langauge one, the specific technical contents of this section are OCaml-specific.
In the case of most other PBT frameworks \cn, the randomness library used operates on by machine
integers, so this \emph{particular} inefficiency does not exist.
